---
title: "Analyzing/Matching GDP data with OECD status for selected countries"
author: "Vishal Ahir"
date: "June 12, 2016"
output: 
  html_document:
    keep_md: true
---

### Purpose
The purpose of this analysis is to read/prep GDP ranking data set of 190 countries for year 2012, merge it with a separate data set that contains details about OECD (Organization for Economic Co-operation and Development) status of countries and provide some answers related to OECD and non-OECD country groups GDP rankings. We will also divide countries based on GDP Rank into 5 groups and see how these groups match to their respective Income Groups.  

OECD - The Organization for Economic Co-operation and Development  
The mission of the Organisation for Economic Co-operation and Development is to promote policies that will improve the economic and social well-being of people around the world. - (**www.oecd.org**)  

Following are the URLs of 2 data sets that will be utilized for this analysis:  
1)  GDP Data : https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv  
2)  Edu Data : https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv  

### Dowloading GDP data set  
  
```{r}
# As a first step, we will store URL for GDP data in a R variable called URL1
URL1 <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"

#Since URL points to a CSV file that has variables separated by comma, we will try and read this data
# into a R data frame using read.csv command

GDPData <- read.csv(URL1, header = FALSE, sep = ",", quote = "\"", fill=TRUE, stringsAsFactors = FALSE)

#As you can see, we are storing the data from dowloaded CSV file into a data frame - GDPData.
#The header = False argument to this function indicates header from CSV file will not be used.
# The header from input file was skipped as there was no meaningful data present.

# Below we are using str() function to check out observations/variables for the downloaded dataset.
str(GDPData)


```
### Dowloading EDU data set

```{r}

# 2nd data set, loaded using same command as first one.
# In this case, header argument is kept true as the file contains meaningful variable header names.

URL2 <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv" 
edu_df <- read.csv(URL2, header=TRUE, sep = ",", quote="\"", fill=TRUE, stringsAsFactors = FALSE)

# Below we are using str() function to check out observations/variables for the downloaded dataset.
str(edu_df)

```

### Prepping GDP data for analysis  
In this block we will prep the GDP data - remove records with missing values, converting data types, sorting, etc.

```{r}
# 
library(plyr)

# From visually inspecting the GDP data set, it is obvious the first 5 rows are of no use as they do not contain
# valid header fields and a lot of extra commas.
# So we will remove first 5 rows of the GDP data set.
GDPData <-GDPData [-(1:5),]

# The second column of GDP Data set contains GDP Rank so we will convert that column to an integer field.
GDPData$V2 <- as.integer(GDPData$V2)


# Now we strip the GDP data set down to only first 5 columns as only these columns are relevant to our analysis.
GDPData2 <- GDPData[,1:5]

#Assigning colume names
names(GDPData2) <- c("CountryCode", "Rank", "Blank", "Economy", "USDollars")

#Below we are removing rows that are missing Rank values for any countries.
GDPData2 <- subset(x = GDPData2, !is.na(GDPData2$Rank))

# The fifth column of GDP data set contains actual GDP in terms of millions of dollars for respective country.
# However the format of that field is XXX,XXX,XXX. So in order to convert this field to numeric we need
# remove those commas and ensure only numbers remain.
GDPData2$USDollars <- gsub(",", "", GDPData2$USDollars)

# Now converting USDollars to numeric field.
GDPData2$USDollars <- as.numeric(GDPData2$USDollars)

# Sorting GDPData2 DF using USDollars.
GDPData2 <- arrange(GDPData2, USDollars)

str(GDPData2)

summary(GDPData2$USDollars)

```

### Prepping EDU data set for analysis  
In this block we will prep the EDU data - remove records with missing values, converting data types, sorting, etc.

``` {r}

#Taking only the first five columns from the edu dataset as only those are relevant.
edu_df_5 <- edu_df[ , 1:5]

#Checking all unique values and their counts using table func.
table(edu_df_5$Income.Group)


```

*Outcome of the table function call states we have 24 blanks or missing values from the EDU data set.*  


### Merging 2 data sets

```{r}

#Now that we have both data sets ready for analysis, we can merge these datasets using common variable "CountryCode".
join_gdp_edu <- merge(GDPData2, edu_df_5, by="CountryCode", all = T)

# Analyzing how many values are missing for Rank field in combined dataset. We can see we have 45 NAs
summary(join_gdp_edu$Rank)

# Analyzing how many values are missing for USDollars (GDP) field in combined dataset. We can see we have 45 NAs as well.
summary(join_gdp_edu$USDollars)

# Removing records with missing values.
gdp_edu_no_na <- subset(join_gdp_edu, !is.na(Rank) & !is.na(Income.Group))

# Sorting combined data set on USDollars (GDP) ascending.
gdp_edu_no_na <- arrange(gdp_edu_no_na, USDollars)

```

### Questions related to analysis

* Question 1 : Match the data based on the country shortcode. How many of the IDs match?  

``` {r}
# We will match function between GDPData2 and edu_df_5 data frames to see how many rows match
# Rows that dont match will be marked as 0 else we will get the record number for the record that matches.

match_count <- match(GDPData2$CountryCode, edu_df_5$CountryCode, nomatch=0)

# Summing up all the records that match.
cnt <- sum(match_count != 0)

```  
**Answer**  In the combined data set, total match count based on IDs is : `r cnt` 


* Question 2 : Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame?  

``` {r}
# Sorting combined data set on USDollars (GDP) ascending.
gdp_edu_no_na <- arrange(gdp_edu_no_na, USDollars)
thirteen_row <- gdp_edu_no_na[13,]
cntry <- thirteen_row$Long.Name

```  
**Answer** Thirteen country is : `r cntry`

* Question 3 : What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?  

``` {r}
# Take subset out of gdp_edu_no_na DF Rank Variable for only those records where Income Group is High Income OECD.
# This subset or Rank is then passed to mean() function to get average rank of such countries.
avg_rank_oecd <- mean(subset(gdp_edu_no_na$Rank, gdp_edu_no_na$Income.Group == "High income: OECD" ))

# Take subset out of gdp_edu_no_na DF Rank Variable for only those records where Income Group is High Income nonOECD.
# This subset or Rank is then passed to mean() function to get average rank of such countries.
avg_rank_non_oecd <- mean(subset(gdp_edu_no_na$Rank, gdp_edu_no_na$Income.Group == "High income: nonOECD" ))

```

***Answer*** Average GDP Ranking for High Income OECD is : `r avg_rank_oecd`
             Average GDP Ranking for High Income nonOECD is : `r avg_rank_non_oecd`

* Question 4 : Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.  

``` {r}
#We will use qplot to plot GDP for all countries.
library(ggplot2)

# Plotting chart with raw USDollars(GDP) values.
qplot (CountryCode, USDollars, data = gdp_edu_no_na, color=Income.Group, xlab="Country", ylab = "GDP")

#Plotting same chart using log of USDollars.
qplot (CountryCode, log(USDollars), data = gdp_edu_no_na, color=Income.Group, xlab="Country", ylab = "GDP")

```

* Question 5 : Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?  

```{r}

# Breaking 189 ranks into 5 groups.
attach(gdp_edu_no_na)
gdp_edu_no_na$rank_groups[Rank <= 38] <- 1
gdp_edu_no_na$rank_groups[Rank > 38 & Rank <= 76] <- 2
gdp_edu_no_na$rank_groups[Rank > 76 & Rank <= 114] <- 3
gdp_edu_no_na$rank_groups[Rank > 114 & Rank <= 152] <- 4
gdp_edu_no_na$rank_groups[Rank >  152] <- 5

# Create a table of rank_groups vs Income Groups to identify Lower Middle Income in nations with 
# highest GDP.

table(gdp_edu_no_na$rank_groups, gdp_edu_no_na$Income.Group)

```
**Answer** After looking at table function call above, we can there are 5 countries that are part of 38 countries with highest GDP and are also part of Lower Middle Income group.  

### Conclusion

  * GDP Dataset
    * 1)  GDP dataset had a total of 326 observations with 10 variables.
    * 2)  Out of these only 190 countries had their GDP listed along with respective GDP Ranks. So 136 rows were NAs.
    * 3)  USA is the ranked as country with highest GDP among all 190 countries.
  * EDU Dataset
    * 1)  EDU dataset had a total  of 234 observations of 31 variables.
    * 2)  When we matched EDU dataset with GDP dataset, we found 189 matches. So there was no match for 45 countries in EDU dataset.
    * 3)  Plot on GDP with log shows equal distribution.
