---
title: "Analyzing/Matching GDP data with OECD status for selected countries"
author: "Vishal Ahir"
date: "June 12, 2016"
output: 
  html_document:
    keep_md: true
---

### Purpose
The purpose of this analysis is to analyze GDP rankings data set of ~190 countries for year 2012, merge it with a separate data set that contains details about OECD status of countries from 1st data set and provide answers related to OECD and non-OECD country groups GDP rankings.  

OECD - The Organization for Economic Co-operation and Development  
The mission of the Organisation for Economic Co-operation and Development (OECD) is to promote policies that will improve the economic and social well-being of people around the world. - (**www.oecd.org**)

Following are the URLs of 2 data sets that will be utilized for this analysis:  
1)  GDP Data : https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv  
2)  Edu Data : https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv  

### Dowloading 2 data sets
  
```{r}
# As a first step, we will store URL for GDP data in a R variable called URL1
URL1 <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"

#Since URL points to a CSV file that has variables separated by comma, we will try and read this data
# into a R data frame using read.csv command

GDPData <- read.csv(URL1, header = FALSE, sep = ",", quote = "\"", fill=TRUE, stringsAsFactors = FALSE)

#As you can see, we are storing the data from dowloaded CSV file into GDPData data frame.
#The header = False argument to this function indicates header from CSV file will not be used.
# The header from input file was skipped as there was no meaningful data present.

# Below we are using str() function to check out observations/variables for the downloaded dataset.
str(GDPData)

# 2nd data set, loaded using same command as first one.
# In this case, header argument is kept true as the file contains meaningful variable header names.

URL2 <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv" 
edu_df <- read.csv(URL2, header=TRUE, sep = ",", quote="\"", fill=TRUE, stringsAsFactors = FALSE)

# Below we are using str() function to check out observations/variables for the downloaded dataset.
str(edu_df)

```

### Prepping data for analysis  
In this block we will prep the data - remove records with missing values, converting data types, sorting, etc.


```{r}
# 
library(plyr)

GDPData$V2 <- as.integer(GDPData$V2)

GDPData2 <- subset(x = GDPData, !is.na(GDPData$V2))

GDPData2 <- arrange(GDPData2, desc(V2))


```

### Merging 2 data sets



### Conclusion